Gravitational Lens Modelling in a Citizen Science Context

### 2014-10-23 text/ms.tex
edits following AV

%\def\half{{\textstyle\frac12}}
% makes it easier to have consistent writing, and easy change if needed
\newcommand{\spl}{SpaghettiLens\xspace}
\newcommand{\sw}{SpaceWarps\xspace}

@@

\author[Küng et al]{Rafael Küng,$^{1}$
Prasenjit Saha,$^{1}$
Anupreeta More,$^{2}$
Elisabeth Baeten,$^{3}$
\newauthor
Jonathan Coles,$^{4}$
Claude Cornen,$^{3}$
Christine Macmillan,$^{3}$
Phil Marshall,$^{5}$ 
\newauthor
Surhud More,$^{2}$
Jonas Odermatt,$^{6}$
Aprajita Verma$^{7}$
and Julianne K. Wilcox$^{3}$
%
\\
%
$^{1}$Physik-Institut, University of Zurich, 190 Winterthurerstrasse, 8057 Zurich, Switzerland\\
$^{2}$Kavli Institute for the Physics and Mathematics of the Universe, University of Tokyo, 5-1-5 Kashiwanoha, Kashiwa-shi 277-8583, Japan\\
$^{3}$Zooniverse, c/o Astrophysics Department, University of Oxford, Oxford \\
$^{4}$Exascale Research Computing Lab, Campus Teratec, 2 Rue de la Piquetterie, 91680 Bruyeres-le-Chatel, France\\
$^{5}$Kavli Institute for Particle Astrophysics and Cosmology, Stanford University, 452 Lomita Mall, Stanford, CA 94035, USA\\
$^{6}$Kantonsschule Zug, L\"ussiweg 24, 6300 Zug, Switzerland\\
$^{7}$Oxford Astrophysics, Keble Road, Oxford OX1 3RH, UK\\
}

@@

We develop a method to enable collaborative modelling of
gravitational lenses and lens candidates, that could be used by
non-professional lens enthusiasts.  It uses an existing free-form
modelling program (GLASS), but enables the input to this code to be
provided in a novel way, via a user-provided diagram that is
essentially a sketch of an arrival-time surface.

An implementation of this method, \spl, is tested in a modelling
challenge that used 29 simulated lenses that had been drawn from the
set created for the \sw citizen-science lens search.  We find that
volunteers from this online community asserted the image parities and
time ordering consistently in some lenses, but made errors in other
lenses depending on the image morphology. While errors in image parity
and time ordering lead to large errors in the mass distribution, the
enclosed mass was found to be more robust: the model-derived Einstein
radii found by the volunteers were consistent with those produced by
one of the professional team, suggesting that given the appropriate
tools, gravitational lens modelling is a data analysis activity that
can be crowd-sourced to good effect. Ideas for improvement are
discussed; these include (a)~overcoming the tendency of the models to
be shallower than the correct answer in test cases, leading to
systematic over-estimation of the Einstein radius by 10\% at present,
and (b)~detailed modelling of arcs.

@@

\begin{keywords}
\end{keywords}

@@

In \secref{Fermat} we introduce a markup system for lensed images,
which we call a ``spaghetti diagram.''  A spaghetti diagram resembles
the visible image system, in a cartoon-like way, and at the same time
it encodes the basis of a mass model.  This supplies an intuitive link
between the image system and the mass distribution, which look
frustratingly different from each other.  Spaghetti diagrams are
essentially the saddle-point contours originally introduced to
gravitational lensing by \cite{1986ApJ...310..568B} as a way of
classifying lensed images.  They are sometimes shown as part of the
output of lens models \citep[for
example][]{2001ApJ...557..594R,2003ApJ...590...39K,Lubini2012}.  In
the present work, however, spaghetti diagrams are the {\em input\/}
through which the modeller tells the program what to do.

@@

There is, however, plenty of room for improvement:
\begin{enumerate}
\item \spl tends to overestimate the Einstein radius (evident from
  \Figref{ER_per_sim}), and the model density profiles tend to be
  shallower than the SIE model used for generating the sims.  The
  likely explanation is that while the sims are steeply peaked at the
  centre, the pixellated mass model fixes a comparatively large area
  near the central at constant density. Allowing smaller pixels in the
  centre region, thus enabling a steeper centre \citep[similar to the
    ``high resolution'' feature implemented in][]{2014arXiv1401.7990C}
  may remove this bias.  The use of simply parameterised,
  appropriately steeply-profiled models would also avoid the problem.
\item Currently, \spl does not attempt to model the source shape; the
  user identifies the brightest points on the image, and these are
  taken as images of a point-like source, whose positions must be
  reproduced exactly. For generating a synthetic image, a conical
  source profile is assumed. Fitting for the source profile to
  optimize resemblance to the observed lensed image after the lens
  model has been generated, is algorithmically straightforward
  \citep[cf.][]{2003ApJ...590..673W,2006MNRAS.371..983S} and planned
  to be implemented.  This would alleviate another problem with \spl,
  which is that there is as yet no quantitative figure of merit for
  any given model: assessment of each model is a judgment call based
  on the synthetic image, and on whether the mass distribution and the
  arrival-time surface show suspicious features.  Another possibility
  would be use the \spl models as a feeder to a different
  lens-modelling program which already implements source-profile
  fitting.
\item Another limitation so far in \spl is that the lens is assumed to
  be dominated by one galaxy, which puts most galaxy-group lenses
  beyond the reach of the modeler. Since complicated group lenses are
  some of the most interesting candidates present, removing this
  limitation is most desirable.  From the users' point of view, it
  would mean that spaghetti contours with more than one maximum can be
  allowed.  For examples, see Figure 5c in \citep{2001ApJ...557..594R}
  and Figure 4b in \cite{2003ApJ...590...39K}.
\item At present, a single false-color composite is used as the data.
  An option could be added to use all available filters, individually
  or in combination, at the user wishes.
\item As mentioned above, the option of revising an already-archived
  model is already available.  Desired now are tools for comparing
  different models of a given system, both visually and through
  different statistical measures.  As evidenced by a current
  collaborative modelling effort, a particularly interesting candidate
  can lead to an extended discussion and dozens of models, that
  in some way sample the high likelihood region of model parameter
  space.
\item Better tutorial materials are also needed, and this would
  address some of the problem areas found in the modelling challenge.
  For example, we saw in \secref{tests.t1} that volunteers are most
  prone to making errors in two situations: when in identifying an arc-
  like structure while placing the points, and in identifying the
  correct ordering of the points in nearly-symmetric configurations.
  Better and more detailed introductory materials would also allow
  the community of modellers to grow faster and without individual
  instructions by experts or experienced volunteers.
\end{enumerate}

>>>>

%\def\half{{\textstyle\frac12}}
% makes it easier to have consistent writing, and easy change if needed
\newcommand{\spl}{SpaghettiLens\xspace}
\newcommand{\sw}{Space Warps\xspace}

@@

\author[Küng et al]{Rafael Küng,$^{1}$
Prasenjit Saha,$^{1}$
Anupreeta More,$^{2}$
Elisabeth Baeten,$^{3}$
\newauthor
Jonathan Coles,$^{4}$
Claude Cornen,$^{3}$
Christine Macmillan,$^{3}$
Phil Marshall,$^{5}$ 
\newauthor
Surhud More,$^{2}$
Jonas Odermatt,$^{6}$
Aprajita Verma$^{7}$
and Julianne K. Wilcox$^{3}$
%
\\
%
$^{1}$Physik-Institut, University of Zurich, 190 Winterthurerstrasse, 8057 Zurich, Switzerland\\
$^{2}$Kavli Institute for the Physics and Mathematics of the Universe, University of Tokyo, 5-1-5 Kashiwanoha, Kashiwa-shi 277-8583, Japan\\
$^{3}$Zooniverse, c/o Astrophysics Department, University of Oxford, Oxford \\
$^{4}$Exascale Research Computing Lab, Campus Teratec, 2 Rue de la Piquetterie, 91680 Bruyeres-le-Chatel, France\\
$^{5}$Kavli Institute for Particle Astrophysics and Cosmology, Stanford University, 452 Lomita Mall, Stanford, CA 94035, USA\\
$^{6}$Kantonsschule Zug, L\"ussiweg 24, 6300 Zug, Switzerland\\
$^{7}$Sub-department of Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford, OX1 3RH, UK\\
}

@@

We develop a method to enable collaborative modelling of
gravitational lenses and lens candidates, that could be used by
non-professional lens enthusiasts.  It uses an existing free-form
modelling program (GLASS), but enables the input to this code to be
provided in a novel way, via a user-generated diagram that is
essentially a sketch of an arrival-time surface.

We report on an implementation of this method, \spl, that has been
tested in a modelling challenge using 29 simulated lenses drawn from a
larger set created for the \sw citizen science strong lens search. We
find that volunteers from this online community asserted the image
parities and time ordering consistently in some lenses, but made
errors in other lenses depending on the image morphology. While errors
in image parity and time ordering lead to large errors in the mass
distribution, the enclosed mass was found to be more robust: the
model-derived Einstein radii found by the volunteers were consistent
with those produced by one of the professional team, suggesting that
given the appropriate tools, gravitational lens modelling is a data
analysis activity that can be crowd-sourced to good effect. Ideas for
improvement are discussed; these include (a)~overcoming the tendency
of the models to be shallower than the correct answer in test cases,
leading to systematic over-estimation of the Einstein radius by 10\%
at present, and (b)~detailed modelling of arcs.

@@

\begin{keywords}
gravitational lensing: strong, methods: numerical
\end{keywords}

@@

In \secref{Fermat} we introduce a markup system for lensed images,
which we call a ``spaghetti diagram.''  A spaghetti diagram resembles
the visible image system, in a cartoon-like way, and at the same time
it encodes the basis of a mass model.  This supplies an intuitive link
between the image system and the mass distribution, which look
frustratingly different from each other.  Spaghetti diagrams are
essentially the saddle-point contours originally introduced to
gravitational lensing by \cite{1986ApJ...310..568B} as a way of
classifying lensed images.  They are sometimes shown as part of the
output of lens models \citep[for
example][]{2001ApJ...557..594R,2003ApJ...590...39K,Lubini2012}.  In
the present work, however, spaghetti diagrams are the {\em input\/}
through which the modeller tells the \spl program what to do.

@@

There is, however, plenty of room for improvement:
\begin{enumerate}
\item \spl tends to overestimate the Einstein radius (evident from
  \Figref{ER_per_sim}), and the model density profiles tend to be
  shallower than the SIE model used for generating the sims.  The
  likely explanation is that while the sims are steeply peaked at the
  centre, the pixellated mass model fixes a comparatively large area
  near the central at constant density. Allowing smaller pixels in the
  centre region, thus enabling a steeper centre \citep[similar to the
    ``high resolution'' feature implemented in][]{2014arXiv1401.7990C}
  may remove this bias.  The use of simply parameterised,
  appropriately steeply-profiled models would also avoid the problem.
\item Currently, \spl does not attempt to model the source shape; the
  user identifies the brightest points on the image, and these are
  taken as images of a point-like source, whose positions must be
  reproduced exactly. For generating a synthetic image, a conical
  source profile is assumed. Fitting for the source profile to
  optimize resemblance to the observed lensed image after the lens
  model has been generated, is algorithmically straightforward
  \citep[cf.][]{2003ApJ...590..673W,2006MNRAS.371..983S} and planned
  to be implemented.  This would alleviate another problem with \spl,
  which is that there is as yet no quantitative figure of merit for
  any given model: assessment of each model is a judgment call based
  on the synthetic image, and on whether the mass distribution and the
  arrival-time surface show suspicious features.  Another possibility
  would be use the \spl models as a feeder to a different
  lens-modelling program that already implements source-profile
  fitting.
\item Another limitation so far in \spl is that the lens is assumed to
  be dominated by one galaxy, which puts most galaxy-group lenses
  beyond the reach of the modeler. Since complicated group lenses are
  some of the most interesting candidates present, removing this
  limitation is most desirable.  From the users' point of view, it
  would mean that spaghetti contours with more than one maximum can be
  allowed.  For examples, see Figure 5c in \citep{2001ApJ...557..594R}
  and Figure 4b in \cite{2003ApJ...590...39K}.
\item At present, a single false-color composite is used as the data.
  An option could be added to use all available filters, individually
  or in combination, at the user wishes.
\item As mentioned above, the option of revising an already-archived
  model is already available.  Desired now are tools for comparing
  different models of a given system, both visually and through
  different statistical measures.  As evidenced by a current
  collaborative modelling effort, a particularly interesting candidate
  can lead to an extended discussion and dozens of models, that
  in some way sample the high likelihood region of model parameter
  space.
\item Better tutorial materials are also needed, and this would
  address some of the problem areas found in the modelling challenge.
  For example, we saw in \secref{tests.t1} that volunteers are most
  prone to making errors in two situations: when in identifying an arc-
  like structure while placing the points, and in identifying the
  correct ordering of the points in nearly-symmetric configurations.
  Better and more detailed introductory materials would also allow
  the community of modellers to grow faster and without individual
  instructions by experts or experienced volunteers.
\end{enumerate}
